---
title: "Carbon Dioxide and Temperature Over Time"
execute:
  echo: false
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(janitor)
library(scales)
library(rstanarm)
```


```{r read initial data}
joined = read_csv("data/joined.csv")
```


``` {r c02 by year}
joined |>
  summarize("carbon_dioxide_emissions_sum" = sum(carbon_dioxide_emissions, na.rm = TRUE),
            .by=year) |>
  ggplot(aes(x=year,
             y=carbon_dioxide_emissions_sum,
             group=1)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  scale_x_continuous(limits = c(1990, 2018), breaks = seq(from=1990, to=2018, by=1)) +
  scale_y_continuous(limits = c(25000, 40000),
                     breaks = seq(20000, 40000, 2500),
                     labels = scales::comma_format()) +
  labs(x="Year",
       y="Carbon Dioxide Emissions (MtC02e)",
       title="Carbon Dioxide Emissions As Time Goes On",
       subtitle="carbon dioxide increases rapidly as time goes on")
```

## Summary

Carbon dioxide is increasing the global temperature. The data I used were pulled from various different data sets, including carbon dioxide emissions, temperature change, land temperature, and population on kaggle. I grouped them together by country and year. I also shortened them to only 1990 to 2018. 

## Model Summaries

- Bayesian regression model with carbon dioxide emissions as the dependent variable and year as the independent variable
- Bayesian regression model with average land temperature as the dependent variable and year as the independent variable

## Takeaway
Around 53,000 (+/- 2000) metric tons of carbon dioxide is going to be in the year 2025 and the average land temperature is going to be 20.6 C (+/- 0.6). 